Linear stability of magnetized massive protoplanetary disks

### 2014-03-17 paper/linear.tex
marginal notes
tex updates

\subsection{Boundary conditions}
At the vertical disk boundaries $z=\pm\zmax$ we assume  
$\dvz=\dbx=\dby=0$, and 
\begin{align}
  \dphi^\prime(\zmax) \pm k_x\dphi(\zmax) = 0
\end{align}
for the potential perturbation \citep[see][]{goldreich65a}, where the
upper (lower) sign applies to the upper (lower) disk surface. 
For Case 1 (purely vertical field) we additionally assume $\bm{U}$ is
an even function of $z$, so that $d\bm{U}/dz=0$ 
at $z=0$.

@@

\subsection{Numerical procedure}
We use a pseudo-spectral method to solve the set of linearized
equations. Let
\begin{align}\label{cheby_expand}
  \bm{U}(z) 
  = \sum_{k=1}^{N_z} \bm{U}_k\psi_k(z/\zmax), 
\end{align}
where 
\begin{align}
  \psi_k  = 
  \begin{cases}
    T_{2(k-1)} & \text{Vertical field (case 1)},\\
    T_{k-1}   & \text{Tilted field (case 2)},
  \end{cases}
\end{align}
and  $T_l$ is a Chebyshev polynomial of the first kind of order $l$
\citep{stegun65}. %Note that in case 1
%The number of basis functions is
%$N_z$ and $l_\mathrm{max}=2(N_Z-1)$ is the highest polynomial order. 
For a vertical field we only use even Chebyshev polynomials 
to satisfy the symmetry condition at $z=0$, so in case 1 we only
consider $z\geq 0$.

The pseudo-spectral coefficients $\bm{U}_n$  are obtained by demanding
the set of linear equations to be satisfied at $N_z$ collocation
points along the vertical direction. We choose these points as the
extrema of $T_{l_\mathrm{max}}$ plus end points, where
$l_\mathrm{max}$ is the highest polynomial order.

@@

\subsubsection{Non-dimensionalization}\label{non-dim}
In practice we solve the linearized equations in non-dimensional form,
by defining
%We further write 
\begin{align}
  &z=\hat{z}H,\quad k_x =  \hat{k}_x/H,\quad \sigma = \hat{\sigma}\Omega,
  \quad \delta\bm{v} = \csmid 
  \delta\hat{\bm{v}}, \\ 
  &\delta\bm{B} = B_z\delta\hat{\bm{B}},\quad
  \delta\rho = \rho\hat{W}/\hat{c}_s^2,\quad \delta\Phi =
  \csmid^2\delta\hat{\Phi},  
\end{align} 
where $\hat{c}_s=c_s/\csmid$. We also non-dimensionalize other
background quantities, i.e. $\hat{v}_A=v_A/\csmid$,
$\hat{S}=S/\Omega$, $\hat{\kappa}=\kappa/\Omega$,
$\hat{\Omega}_z=\Omega_z/\Omega$ and $\hat{\eta} = \eta/(H^2\Omega)$.

\subsection{Diagnostics}
We visualize results in terms of dimensionless energy densities. We
define 
\begin{align}
  &E_m \equiv \frac{|\delta \hat{\bm{B}}|^2}{2\beta},\\
  &E_g =
  \frac{\hat{\rho}}{\hat{c}_s^2}\left|\real\left(\hat{W}\dd\hat{\Phi}^*\right)\right|,\\  
  &E_k = \frac{1}{2}\hat{\rho}|\delta\hat{\bm{v}}|^2,\\
  &E_t = \frac{\hat{\rho}|\hat{W}|^2}{2\hat{c}_s^2},
\end{align}
as the perturbed magnetic, gravitational, kinetic and thermal
energies, respectively, which are functions of $z$. The total energy
is $E=E_m+E_g+E_k+E_t$ and we then normalize $E_i$ by
$\mathrm{max}(E)$.

>>>>

\subsection{Domain and boundary conditions}

\subsubsection{Case 1: vertical field}
We impose $\bm{U}$ to be an even function of $z$, so that
$d\bm{U}/dz=0$ at $z=0$. This reduces the computational domain to
$z\in[0,Z_s]$ and permits higher numerical resolution. At the disk
surface we set %$\dbx=\dby=\dvz=0$ 
\begin{align}
  \dbx(Z_s) = \dby(Z_s) = \dvz (Z_s) = 0, 
\end{align}
so the field remains vertical. The potential boundary
condition is 
\begin{align}
  \dphi^\prime(\zmax) + k_x\dphi(\zmax) = 0 \label{pot_bc1}
\end{align}
\citep{goldreich65a}.

\subsubsection{Case 2: tilted field} 
In this more general setup the computational domain is
$z\in[-Z_s, Z_s]$. At the disk surfaces we adopt the `halo' model of
\cite{gammie94}, so that
\begin{align}
  \Delta(\pm Z_s) & = 0, \\
  \dby(\pm Z_s) & = 0,\\
  \dbz(\pm Z_s) \mp \imgi \dbx(\pm Z_s) &= 0,
\end{align}
and the potential boundary condition is 
\begin{align}
  \dphi^\prime(\pm \zmax) \pm k_x\dphi(\pm\zmax) =  - \left. \frac{\Omega^2
    \rho \xi_z}{\rho_0Q}\right|_{\pm\zmax}, \label{pot_bc2}
\end{align}
where $\xi_z=\dvz/\imgi\sigma$ is the vertical Lagrangian
displacement. In practice,  Eq. \ref{pot_bc1} and \ref{pot_bc2} are
very similar because $\rho(\pm\zmax)\ll\rho_0$.

%We adopt two kinds of vertical boundary conditions depending on the
%field geometry. 
%In both case 1 and case 2 the potential boundary
%condition is 
%\begin{align}
%  \dphi^\prime(\zmax) \pm k_x\dphi(\zmax) = 0
%\end{align}

%At the vertical disk boundaries $z=\pm\zmax$ we assume  
%$\dvz=\dbx=\dby=0$, and

%for the potential perturbation \citep[see][]{goldreich65a}, where the
%upper (lower) sign applies to the upper (lower) disk surface. 
%For Case 1 (purely vertical field) we additionally assume $\bm{U}$ is
%an even function of $z$, so that $d\bm{U}/dz=0$ 
%at $z=0$.

@@

\subsection{Numerical procedure}
We use a pseudo-spectral method to solve the set of linearized
equations. Let
\begin{align}\label{cheby_expand}
  \bm{U}(z) 
  = \sum_{k=1}^{N_z} \bm{U}_k\psi_k(z/\zmax), 
\end{align}
where 
\begin{align}
  \psi_k  = 
  \begin{cases}
    T_{2(k-1)} & B_y \equiv 0 \text{ (case 1)},\\
    T_{k-1}   & B_y\neq 0 \text{ (case 2)},
  \end{cases}
\end{align}
and  $T_l$ is a Chebyshev polynomial of the first kind of order $l$
\citep{stegun65}. Note that for case 1 the midplane symmetry condition
is taken care of by the choice of basis functions. 
%Note that in case 1
%The number of basis functions is
%$N_z$ and $l_\mathrm{max}=2(N_Z-1)$ is the highest polynomial order. 
%For a vertical field we only use even Chebyshev polynomials 
%to satisfy the symmetry condition at $z=0$, so in case 1 we only
%consider $z\geq 0$.

The pseudo-spectral coefficients $\bm{U}_n$  are obtained by demanding
the set of linear equations to be satisfied at $N_z$ collocation
points along the vertical direction. We choose these points as the
extrema of $T_{l_\mathrm{max}}$ plus end points, where
$l_\mathrm{max}$ is the highest polynomial order. Our standard
resolution is $N_z=256$.

@@

\subsubsection{Non-dimensionalization}\label{non-dim}
We solve the linearized equations in non-dimensional form,
by defining
%We further write 
\begin{align}
  &z=\hat{z}H,\quad k_x =  \hat{k}_x/H,\quad \sigma = \hat{\sigma}\Omega,
  \quad \delta\bm{v} = \csmid 
  \delta\hat{\bm{v}}, \\ 
  &\delta\bm{B} = B_z\delta\hat{\bm{B}},\quad
  \delta\rho = \rho\hat{W}/\hat{c}_s^2,\quad \delta\Phi =
  \csmid^2\delta\hat{\Phi},  
\end{align} 
where $\hat{c}_s=c_s/\csmid$. We also non-dimensionalize other
background quantities, i.e. $\hat{v}_A=v_A/\csmid$,
$\hat{S}=S/\Omega$, $\hat{\kappa}=\kappa/\Omega$,
$\hat{\Omega}_z=\Omega_z/\Omega$ and $\hat{\eta} = \eta/(H^2\Omega)$.

\subsection{Diagnostics}
We visualize results in terms of dimensionless energy densities. We
define 
\begin{align}
  &E_m \equiv \frac{|\delta \hat{\bm{B}}|^2}{2\beta},\\
  &E_g =
  \frac{\hat{\rho}}{\hat{c}_s^2}\left|\real\left(\hat{W}\dd\hat{\Phi}^*\right)\right|,\\  
  &E_k = \frac{1}{2}\hat{\rho}|\delta\hat{\bm{v}}|^2,\\
  &E_t = \frac{\hat{\rho}|\hat{W}|^2}{2\hat{c}_s^2},
\end{align}
as the perturbed magnetic, gravitational, kinetic and thermal
energies, respectively, which are functions of $z$. The total energy
is $E=E_m+E_g+E_k+E_t$. We use $\avg{\cdot}$ to denote an average over
$z$. 
%and we then normalize $E_i$ by
%$\mathrm{max}(E)$.

### 2014-03-17 paper/paper.tex
minor paper update
paper update
major update to tex
marginal notes
tex updates

%\documentclass[12pt]{aastex}
\documentclass[iop]{emulateapj}
%\documentclass{article}
\usepackage{draftwatermark}
\usepackage{paralist}
\shorttitle{MRI in massive disks}
\shortauthors{M-K. Lin}

\usepackage{amsmath}
%\usepackage{url}
\usepackage{bm}
%\usepackage[font=small,format=plain,labelfont=bf,up,textfont=it,up]{caption}
\newcommand{\p}{\partial}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\sbar}{\bar{\sigma}} 
\newcommand{\imgi}{\mathrm{i}} 
\newcommand{\ii}{\mathrm{i}}
\newcommand{\C}{\mathcal{C}^\lambda}
\newcommand{\wop}[1]{\mathcal{V}^{(#1)}}
\newcommand{\avg}[1]{\langle{#1}\rangle}
\newcommand{\zhat}{\hat{z}}
\newcommand{\rhat}{\hat{r}}
\newcommand{\he}{\mathcal{H}}
\newcommand{\amp}{\mathcal{A}}
\newcommand{\dd}{\delta}
\newcommand{\real}{\operatorname{Re}}
\newcommand{\imag}{\operatorname{Im}}
\newcommand{\sgn}{\operatorname{sgn}}
\newcommand{\zmax}{Z_s}
\newcommand{\wbar}{\tilde{W}}
\newcommand{\qbar}{\tilde{Q}}
\newcommand{\dbigH}{\frac{H^\prime}{H}}
\newcommand{\Rp}{\frac{\rho_0^\prime}{\rho_0}}
\newcommand{\Fp}{\frac{g^\prime}{g}}
\newcommand{\Rpp}{\frac{\rho_0^{\prime\prime}}{\rho_0}}
\newcommand{\Fpp}{\frac{g^{\prime\prime}}{g}}
\newcommand{\ils}{L_s^{-1}}
\newcommand{\ilp}{L_p^{-1}}
\newcommand{\ihp}{H_p^{-1}}
\newcommand{\ihs}{H_s^{-1}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\csmid}{c_{s0}} 
\newcommand{\hiso}{h_\mathrm{iso}}
\newcommand{\Hiso}{H_\mathrm{iso}}
\newcommand{\zeus}{\texttt{ZEUS-MP }}
\newcommand{\rsph}{r_\mathrm{sph}}
\newcommand{\dvx}{\dd v_x}
\newcommand{\dvy}{\dd v_y}
\newcommand{\dvz}{\dd v_z}
\newcommand{\dbx}{\dd B_x}
\newcommand{\dby}{\dd B_y}
\newcommand{\dbz}{\dd B_z}
\newcommand{\w}{ \tilde{W}}
\newcommand{\dphi}{\dd \Phi}
\defcitealias{lin12}{L12}

@@

 \title{Resistive magneto-rotational instability in massive
  protoplanetary disks}

\author{Min-Kai Lin}
\affil{Canadian Institute for Theoretical Astrophysics,
  60 St. George Street, Toronto, ON, M5S 3H8, Canada}
\email{mklin924@cita.utoronto.ca}

@@

\section{Introduction}

\input{setup}
\input{linear}
%\input{results}
%\input{iso}
%\input{summary}

%% \acknowledgments
%% {\bf
%% I thank the referee, P. Barge, for suggesting the idea of a tilted vorticity column. }
%% I also thank S.-J. Paardekooper for comments on the first version of this paper.

>>>>

%\documentclass[12pt]{aastex}
\documentclass[iop]{emulateapj}
%\documentclass{article}
%\usepackage{draftwatermark}
\usepackage{paralist}
\shorttitle{Magnetized massive disks}
\shortauthors{M-K. Lin}

\usepackage{amsmath}
%\usepackage{url}
\usepackage{bm}
%\usepackage[font=small,format=plain,labelfont=bf,up,textfont=it,up]{caption}
\newcommand{\p}{\partial}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\sbar}{\bar{\sigma}} 
\newcommand{\imgi}{\mathrm{i}} 
\newcommand{\ii}{\mathrm{i}}
\newcommand{\C}{\mathcal{C}^\lambda}
\newcommand{\wop}[1]{\mathcal{V}^{(#1)}}
\newcommand{\avg}[1]{\langle{#1}\rangle}
\newcommand{\zhat}{\hat{z}}
\newcommand{\rhat}{\hat{r}}
\newcommand{\he}{\mathcal{H}}
\newcommand{\amp}{\mathcal{A}}
\newcommand{\dd}{\delta}
\newcommand{\real}{\operatorname{Re}}
\newcommand{\imag}{\operatorname{Im}}
\newcommand{\sgn}{\operatorname{sgn}}
\newcommand{\zmax}{Z_s}
\newcommand{\wbar}{\tilde{W}}
\newcommand{\qbar}{\tilde{Q}}
\newcommand{\dbigH}{\frac{H^\prime}{H}}
\newcommand{\Rp}{\frac{\rho_0^\prime}{\rho_0}}
\newcommand{\Fp}{\frac{g^\prime}{g}}
\newcommand{\Rpp}{\frac{\rho_0^{\prime\prime}}{\rho_0}}
\newcommand{\Fpp}{\frac{g^{\prime\prime}}{g}}
\newcommand{\ils}{L_s^{-1}}
\newcommand{\ilp}{L_p^{-1}}
\newcommand{\ihp}{H_p^{-1}}
\newcommand{\ihs}{H_s^{-1}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\csmid}{c_{s0}} 
\newcommand{\hiso}{h_\mathrm{iso}}
\newcommand{\Hiso}{H_\mathrm{iso}}
\newcommand{\zeus}{\texttt{ZEUS-MP }}
\newcommand{\rsph}{r_\mathrm{sph}}
\newcommand{\dvx}{\dd v_x}
\newcommand{\dvy}{\dd v_y}
\newcommand{\dvz}{\dd v_z}
\newcommand{\dbx}{\dd B_x}
\newcommand{\dby}{\dd B_y}
\newcommand{\dbz}{\dd B_z}
\newcommand{\w}{ \widetilde{W}}
%\newcommand{\sbar}{\bar{\sigma}}
\newcommand{\dphi}{\dd \Phi}
\defcitealias{lin12}{L12}

@@

\title{Linear stability of magnetized massive protoplanetary disks}

\author{Min-Kai Lin (+Menou, Fromang, Youdin)}
\affil{Canadian Institute for Theoretical Astrophysics,
  60 St. George Street, Toronto, ON, M5S 3H8, Canada}
\email{mklin924@cita.utoronto.ca}

@@

\section{Introduction}
Understanding the evolution of protoplanetatry disks (PPDs) is key to planet
formation theory. A central aspect of PPD evolution is the mechanism
of angular momentum transport.

\input{setup}
\input{linear}
\input{results}
\input{results_gi}
\input{results_azi}

### 2014-03-17 paper/results.tex
tex updates

\section{Results: MRI in self-gravitating disks}
In this section we focus on the MRI. We use the vertical field setup
of case 1. The numerical resolution is $N_z=256$.

@@

Here we consider polytropic disks, which have a well-defined
thickness. Numerically, the upper disk boundary is set to
$Z_s=0.99H$. We fix $\beta=100$ and $k_xH=0.1$.

@@

\subsubsection{Layered
  resistivity} 
Here we consider disks with mid-plane Elsasser number $\Lambda_0=0.1$
and a variable resistivity profile with
$A=10^2$. Fig. \ref{poly_layer} compares the magnetic  
energies for $Q=0.2,\,1$ and $4$. They have similar growth rates, $\gamma/\Omega
= 0.53,\,0.64$ and $0.66$, respectively. In the non-self-gravitating
limit ($Q=4$), the MRI is effectively surpressed for
$z\lesssim0.5H$. This is consistent with layered accretion. However,
in the massive disk ($Q=0.2$) the mode occupies a wider vertical
extent because its wavelength (in units of $H$) is larger. This
suggests that in massive disks the MRI is not well localized to the
active layer, as the instability takes on a more global character
compared to non-self-gravitating disks.        
%d is not as well localized

@@

@@

\subsubsection{Ideal case}
We first consider the limit of ideal MHD by adopting a uniform
resistivity with $\Lambda_0=100$. %Non-zero density perturbations
%require $k_x\neq0$ so we consider linear modes as a function of $k_x$.   
Fig. \ref{gravity_energy} shows the maximum gravitational potential
energy perturbation, as a function of $k_x$ for several values of
$\beta$. The potential perturbation is negligible for all cases when 
$k_xH\lesssim 0.5$, since the MRI becomes incompressible as $k_x\to
0$.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_energy_ideal}
  \caption{Maximum gravitational potential energy perturbation
    associated with the MRI in a self-gravitating disk with $Q=0.2$ in
    the limit of ideal MHD ($\Lambda_0=10^2$), for a range of field
    strengths $\beta$.  
    \label{gravity_energy}}
\end{figure}

@@

\subsubsection{Resistive case}
We repeat the above calculation for resistive disks, but fix
$\beta=100$. Fig. \ref{gravity_energy_resis} shows the maximum
perturbed gravitational energy increases with 
resistivity. Interestingly, in the highly resistive case
the MRI can be energeically dominated by self-gravity. An example is shown in
Fig. \ref{mri_massive_resis}, which  plots the energies assocaited with the 
linear MRI mode with $k_xH\sim1.3$ in the $\Lambda_0=0.1$  
disk. The gravitational potential energy significantly exceeds the
magnetic energy near the midplane ($z\lesssim0.3H$).

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_energy_resis}
  \caption{Maximum gravitational potential energy perturbation
    associated with the MRI in a self-gravitating disk with $Q=0.2$,
    for a range of midplane Elsasser numbers $\Lambda_0$ at fixed
    field strength $\beta=100$.    
    \label{gravity_energy_resis}}
\end{figure}

We also calculated the mode in Fig. \ref{mri_massive_resis} with the \emph{Cowling 
  approximation}, where the Poisson equation is ignored in the 
linearized equations and the potential perturbation set to zero  
(formally, we set $Q\to\infty$ in Eq. \ref{lin_poisson}).  
We compare the magnetic energy in the Cowling approximation to that of
the full problem in Fig. \ref{mri_massive_cowling}. The growth rate
increases when self-gravity is included in the linear response, as
expected, but only slightly. $\gamma$ and $E_m$ are similar in both cases,
indicating that the instability in the self-gravitating calculation is
still due to the MRI.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/result_resis_sg}
  \caption{Example of a linear resistive MRI mode with significant
    gravitational potential energy perturbation.  The vertical line
    inidicates $\Lambda=1$.  
    \label{mri_massive_resis}}
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_result_cowling}
  \caption{Magnetic energy associated with the linear mode shown in
    Fig. \ref{mri_massive_resis} (solid) compared with that computed
    under the Cowling approximation (dotted). 
    \label{mri_massive_cowling}}
\end{figure}

@@

We now include self-gravity. In general, the compressible MRI has
$W^{(0)}\neq0$. The Poisson equation implies
an associated potential perturbation,    
\begin{align} 
  \dphi = -\frac{\Omega^2}{c_s^2 Q k^2} W^{(0)}.
\end{align}
Physically, we expect $k^2\geq0$, so that a positive (negative) density
perturbation causes a negative (positive) potential perturbation. We
then regard $\dphi$ to alter the Cowling solution, so
that $\dvx^{0} \to \dvx^{0} + \dvx^{(1)}$ and similarly for $\dvy$ and
$W$. Then 
\begin{align}
  &   k_x\sigma \dphi = v_A^2k^2\dvx^{(1)} + \imgi\sigma\left[\imgi\sigma
  \dvx^{(1)} - 2\Omega\dvy^{(1)} + \imgi k_x W^{(1)}\right], \label{simp_pert1}\\ 
  &  0= v_A^2k_z^2\left[\dvy^{(1)} + \frac{\imgi S
    }{\sigma}\dvx^{(1)}\right] + \imgi\sigma\left[\imgi\sigma\dvy^{(1)} +
  \frac{\kappa^2}{2\Omega}\dvx^{(1)}\right],\label{simp_pert2}\\
  & k_z^2\dphi  = \left(\frac{\sigma^2}{c_s^2}-k_z^2\right)W^{(1)} +
  \sigma k_x \dvx^{(1)}.\label{simp_pert3} 
\end{align}
Now, if the perturbations to the magnetic field remain
unchanged, i.e. the mode remains close to the standard MRI, then
$\dvx^{(1)} \sim 0$ and $\dvy^{(1)}\sim0$, so Eq. \ref{simp_pert2} is
satisfied. Eq. \ref{simp_pert1} then require $\dphi +
W^{(1)}\sim0$. This is compatible with Eq. \ref{simp_pert3} if 
\begin{align}
  \left|k_z^2\right| \gg \left|\frac{\sigma^2}{c_s^2}\right|. \label{cond}
\end{align}
For the ideal MRI,  we assume $k_z\sim \Omega/v_A$. Then
$|\sigma^2/c_s^2k_z^2|\sim |\sigma^2/\Omega^2\beta|\ll1$ because
$|\sigma|\lesssim \Omega$ and we are considering $\beta\gtrsim
10$. Thus Eq. \ref{cond} is generally satisfied.

>>>>

\section{MRI in self-gravitating disks}
In this section we focus on the MRI. We use the vertical field setup
of case 1. We first consider MRI modes with negligble densit/potential
perturbations to see the effect of self-gravity on the MRI through the
background stratification, then go on to consider MRI modes with
significant density/potential perturbations in massive disks.

@@

Here we use polytropic disks, which have a well-defined
thickness. The upper disk boundary is set to 
$Z_s=0.99H$. We fix $\beta=100$ and $k_xH=0.1$ unless otherwise
stated.

@@

\subsubsection{Layered
  resistivity} 
Here we consider disks with mid-plane Elsasser number $\Lambda_0=0.1$
and a variable resistivity profile with
$A=10^2$. Fig. \ref{poly_layer} compares the magnetic  
energies for $Q=0.2,\,1$ and $4$. They have similar growth rates, $\gamma/\Omega
= 0.53,\,0.64$ and $0.66$, respectively. In the non-self-gravitating
limit ($Q=4$), the MRI is effectively surpressed for
$z\lesssim0.5H$. This is consistent with layered accretion. However,
in the massive disk ($Q=0.2$) the mode occupies a wider vertical
extent because its wavelength (in units of $H$) is larger. This
suggests that in massive disks the MRI is not well localized to the
active layer.

%MRI penetrates 
%d is not as well localized

@@

\subsubsection{Dependence on $k_x$}
The above experiments show that with increasing 
disk self-gravity, the MRI becomes more global in the vertical
direction. We find a similar result in the horizontal direction. 
In Fig. \ref{compare_growth_poly_kx} we plot MRI growth rates as a
function of $k_x$ for a range of $Q$ values. Increasing self-gravity
decreases the cut-off horizontal wavenumber for the MRI, implying that
small-scale MRI does not operate in massive disks.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_growth_poly_varQ_kx}
  \caption{MRI growth rates in self-gravitating polytropic disks, as a
    function of the horizontal wavenumber $k_x$. The disk is ideal
    ($\Lambda_0=10^2,\, A=1$) with $\beta = 40$. These modes have negligible
    density/potential perturbations.  
    \label{compare_growth_poly_kx}}
\end{figure}

%these modes are still non-sg
%use beta = 40 so only a small range of kx need to be simulated to
%demonstrate this effect
%{\bf cut-off $k_xH$ decreases with increasing SG -> small-scale MRI
%  stabilized by self-gravity}

@@

\subsubsection{Ideal case}
We first consider the limit of ideal MHD by adopting a uniform
resistivity with $\Lambda_0=100$. %Non-zero density perturbations
%require $k_x\neq0$ so we consider linear modes as a function of $k_x$.   
Fig. \ref{gravity_energy} shows the mean gravitational potential
energy perturbation, as a function of $k_x$ for several values of
$\beta$. The potential perturbation is negligible for all cases when 
$k_xH\lesssim 0.5$, since the MRI becomes incompressible as $k_x\to
0$.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_energy_ideal}
  \caption{Average gravitational potential energy perturbation
    associated with the MRI, compared to the average magnetic energy
    perturbation, in a self-gravitating disk with $Q=0.2$ in 
    the limit of ideal MHD ($\Lambda_0=10^2$), for a range of field
    strengths $\beta$.  
    \label{gravity_energy}}
\end{figure}

@@

\subsubsection{Resistive case}
We repeat the above calculation for resistive disks, but fix
$\beta=100$. Fig. \ref{gravity_energy_resis} shows the 
perturbed gravitational energy increases with   
resistivity. Interestingly, in the highly resistive case
the MRI can be energeically dominated by self-gravity, i.e. the case
with $\Lambda_0=0.1$ and $k_xH\simeq1.3$. We also calculated this mode
in the \emph{Cowling  
  approximation}, where the Poisson equation is ignored in the 
linearized equations and the potential perturbation set to zero  
(formally letting $Q\to\infty$ in Eq. \ref{lin_poisson}).
Fig. \ref{mri_massive_cowling} compares the magnetic energy in the
Cowling approximation to that of 
the full problem in . The growth rate increases when self-gravity is
included in the linear response, as expected, but only
slightly. $\gamma$ and $E_m$ are similar in both cases, indicating
that the instability in the self-gravitating calculation is 
still the MRI.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_energy_resis}
  \caption{Average gravitational potential energy perturbation
    associated with the MRI, compared to the average magnetic energy
    perturbation, in a self-gravitating disk with $Q=0.2$,
    for a range of midplane Elsasser numbers $\Lambda_0$ at fixed
    field strength $\beta=100$.    
    \label{gravity_energy_resis}}
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_result_cowling}
  \caption{Magnetic energy associated with the linear mode with
    largest gravitational-to-magnetic energy ratio in
    Fig. \ref{gravity_energy_resis} (solid) compared with that computed
    under the Cowling approximation (dotted). 
    \label{mri_massive_cowling}}
\end{figure}

%An example is shown in
%Fig. \ref{mri_massive_resis}, which  plots the energies assocaited with the 
%linear MRI mode with $k_xH\sim1.3$ in the $\Lambda_0=0.1$  
%disk. The gravitational potential energy significantly exceeds the
%magnetic energy near the midplane ($z\lesssim0.3H$).

%\begin{figure}
%  \includegraphics[width=\linewidth]{figures/result_resis_sg}
%  \caption{Example of a linear resistive MRI mode with significant
%    gravitational potential energy perturbation.  The vertical line
%    inidicates $\Lambda=1$.  
%    \label{mri_massive_resis}}
%\end{figure}

@@

We now include self-gravity. In general, the compressible MRI has
$W^{(0)}\neq0$. The Poisson equation implies
an associated potential perturbation,    
\begin{align} 
  \dphi = -\frac{\Omega^2}{c_s^2 Q k^2} W^{(0)}.
\end{align}
Physically, we expect $k^2\geq0$, so that a positive (negative) density
perturbation causes a negative (positive) potential perturbation. We
then regard $\dphi$ to alter the Cowling solution, so
that $\dvx^{(0)} \to \dvx^{(0)} + \dvx^{(1)}$ and similarly for $\dvy$ and
$W$. Then 
\begin{align}
  &   k_x\sigma \dphi = v_A^2k^2\dvx^{(1)} + \imgi\sigma\left[\imgi\sigma
  \dvx^{(1)} - 2\Omega\dvy^{(1)} + \imgi k_x W^{(1)}\right], \label{simp_pert1}\\ 
  &  0= v_A^2k_z^2\left[\dvy^{(1)} + \frac{\imgi S
    }{\sigma}\dvx^{(1)}\right] + \imgi\sigma\left[\imgi\sigma\dvy^{(1)} +
  \frac{\kappa^2}{2\Omega}\dvx^{(1)}\right],\label{simp_pert2}\\
  & k_z^2\dphi  = \left(\frac{\sigma^2}{c_s^2}-k_z^2\right)W^{(1)} +
  \sigma k_x \dvx^{(1)}.\label{simp_pert3} 
\end{align}
Now, if the perturbations to the magnetic field remain
unchanged, i.e. the mode remains close to the standard MRI, then
$\dvx^{(1)} \sim 0$ and $\dvy^{(1)}\sim0$, so Eq. \ref{simp_pert2} is
satisfied. Eq. \ref{simp_pert1} then require $\dphi +
W^{(1)}\sim0$. This is compatible with Eq. \ref{simp_pert3} if 
\begin{align}
  \left|k_z^2\right| \gg \left|\frac{\sigma^2}{c_s^2}\right|. \label{cond}
\end{align}
For the ideal MRI,  we assume $k_z\sim \Omega/v_A$. Then
$|\sigma^2/c_s^2k_z^2|\sim |\sigma^2/\Omega^2\beta|\ll1$ because
$|\sigma|\lesssim \Omega$ and we are considering $\beta\gtrsim
10$. Thus Eq. \ref{cond} is generally satisfied.

### 2014-03-17 paper/setup.tex
tex updates

We use the Elsasser number $\Lambda$ as a non-dimensional measure of
conductivity,
\begin{align} 
  \Lambda \equiv \frac{v_A^2}{\eta\Omega},
\end{align}
where $v_A \equiv B_z/\sqrt{\mu_0\rho}$ is the vertical Alfven speed. Example
profiles of $\Lambda$ are shown in Fig. \ref{eqm_resis} for the
isothermal disk. The disk may be considered ideal where $\Lambda
\gtrsim 1$. Since $\hat{\rho}(z/H)$ is weakly dependent on  self-gravity, the
ideal region only expands slightly as $Q$ is decreased.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/elsasser_iso}
  \caption{Elsasser number as prescribed in \S\ref{resis_profile} for
    isothermal disks with $Q=0.2$ (solid) 
    and $Q=10$ (dotted). The vertical line
    indicates $\Lambda=1$ in each case. \label{eqm_resis}}
\end{figure}

>>>>

We use the Elsasser number $\Lambda$ as a non-dimensional measure of
conductivity,
\begin{align} 
  \Lambda \equiv \frac{v_A^2}{\eta\Omega},
\end{align}
where $v_A \equiv B_z/\sqrt{\mu_0\rho}$ is the vertical Alfven speed. 
%Example
%profiles of $\Lambda$ are shown in Fig. \ref{eqm_resis} for the
%isothermal disk. 
The disk may be considered ideal where $\Lambda
\gtrsim 1$. Since $\hat{\rho}(z/H)$ is weakly dependent on  self-gravity, the
ideal region only changes slightly as $Q$ is decreased.

@@

%\begin{figure}
%  \includegraphics[width=\linewidth]{figures/elsasser_iso}
%  \caption{Elsasser number as prescribed in \S\ref{resis_profile} for
%    isothermal disks with $Q=0.2$ (solid) 
%    and $Q=10$ (dotted). The vertical line
%    indicates $\Lambda=1$ in each case. \label{eqm_resis}}
%\end{figure}
